<!-- # parid package: Calculation of parameter identifiability -->
<!-- #                indicators for nonlinear mixed effects models. -->
<!-- # -->
<!-- # Copyright (C) 2023 LAP&P Consultants BV -->
<!-- # Contact: info@lapp.nl -->
<!-- #          www.lapp.nl -->
<!-- #          Archimedesweg 31 -->
<!-- #          2333 CM Leiden, The Netherlands -->
<!-- #  -->
<!-- # This file is part of the parid package. -->
<!-- # The parid package is free software: you can redistribute it -->
<!-- # and/or modify it under the terms of the GNU General Public License as -->
<!-- # published by the Free Software Foundation, either version 3 of the -->
<!-- # License, or (at your option) any later version. -->
<!-- # -->
<!-- # The parid package is distributed in the hope that it will be -->
<!-- # useful, but WITHOUT ANY WARRANTY; without even the implied warranty -->
<!-- # of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the -->
<!-- # GNU General Public License for more details. -->
<!-- # -->
<!-- # You should have received a copy of the GNU General Public License -->
<!-- # along with the parid package. -->
<!-- # If not, see <https://www.gnu.org/licenses/>.  -->

---
title: "parid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{parid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(parid)
```

# Introduction
This is an example showing how to use the parid package.
The example defines a nonlinear mixed effects model, and analyzes its parameter identifiability using three methods:

- the Sensitivity Matrix Method (SMM)
- the Fisher Information Matrix Method (FIMM)
- the Aliasing Method

This example assumes familiarity with mixed effects modeling.
More details on the methods and on parameter identifiability in general can be found [here](https://lapp.nl/lapp-software/parid.html).
R code running this same example can be found in the package, at tests/example.R.

# Model definition
The model is a one-compartmental linear PK model with absorption and a single oral dose.
The model parameters are clearance (CL), volume (V), the absorption rate constant (Ka) and optionally the bio-availability (F); they are defined in terms of underlying structural and inter-individual parameters.
In this example, inter-individual variability is included on clearance.
The output is the concentration, which is modeled with an additive error.

The expected outcome is that this model will not be identifiable because the bio-availability cannot be determined independently from the clearance and volume. If the bio-availability is fixed, then the model should be identifiable.

The next lines of code create four functions defining this model:

- `model`: returns the differential equations describing the time evolution of the model. The input arguments are the time `t`, the current state `y` (i.e., the amounts in the depot and central compartments), and the model parameters `p`.
- `p`: returns the model parameters as function of the structural parameters `theta` and inter-individual parameters `eta`. The dose is also included here. In general, any covariates could be included. For all parameters except CL, the value is simply the value of the corresponding structural parameter.
- `init`: returns the initial values of the state variables, as function of the model parameters `p`. In this example, the depot is initialized to the bio-available fraction of the dose, and the central compartment to zero.
- `output`: returns the model output, that is the concentration (with residual error), as function of the state `y`, the model parameters `p` and the residual error `eps`.

```{r}
model      <- function(t, y, p) { c(-p[["Ka"]]*y[["x1"]], p[["Ka"]]*y[["x1"]] - p[["CL"]]/p[["V"]] * y[["x2"]]) }
p          <- function(theta, eta) { c(Ka = theta[["TVKa"]], CL = theta[["TVCL"]] * exp(eta[["iCL"]]),
                                       V = theta[["TVV"]], F = theta[["TVF"]], Dose = theta[["TVDose"]])  }
init       <- function(p) { c("x1" = p[["F"]] * p[["Dose"]], "x2" = 0) }
output     <- function(y, p, eps) { y[["x2"]] / p[["V"]] + eps[["addErr"]] }
```

# Parameter definition
The next piece of code sets the values for the structural parameters (`theta`).
Some parameters can be considered as fixed.
This is typically done for covariates (in this example the dose).
The vectors `vartheta1` and `vartheta2` define two variants; the first one also sets the bio-availability F as fixed, the second one keeps it variable.

Variance matrices `omega` and `sigma` are defined for the random variables `eta` and `eps`, respectively.
Their row and column names should be the same as the ones used in the model definition.
These matrices are the random-effects parameters.

```{r}
theta      <- c("TVKa" = 1, "TVCL" = 0.2, "TVV" = 2, "TVF" = 0.7, "TVDose" = 10)
vartheta1  <- setdiff(names(theta), c("TVDose", "TVF"))     # Bioavailability F is fixed
vartheta2  <- setdiff(names(theta), "TVDose")               # Bioavailability F is variable
omega      <- diag(0.3, nrow = 1)
colnames(omega) <- row.names(omega) <- c("iCL")
sigma      <- diag(0.1, nrow = 1)
colnames(sigma) <- row.names(sigma) <- c("addErr")
```

# Sampling times
The vector `times` contains the sampling times.
The initialization (in this example, the time of dosing) is always set to zero.

```{r}
times      <- seq(0, 10, 1)
```

# Variational equations
As a first step in the calculation, the variational equations are solved.
This is done two times, once for every setting of fixed parameters as defined above, and will create data frames `vareq1` and `vareq2` containing the derivatives of the model output (the concentration) with respect to the structural and random-effects parameters, evaluated at the sample times. The derivatives are plotted over time.
The plot functions have options controlling which derivatives are plotted. See their documentation for details.

```{r}
vareq1 <- calcVariationsFim(model = model, p = p, init = init, output = output, times = times, symbolic = TRUE,
                           theta = theta, nmeta = colnames(omega), nmeps = colnames(sigma), vartheta = vartheta1)
plotVariations(vareq1)
vareq2 <- calcVariationsFim(model = model, p = p, init = init, output = output, times = times, symbolic = TRUE,
                            theta = theta, nmeta = colnames(omega), nmeps = colnames(sigma), vartheta = vartheta2)
plotVariations(vareq2)
```

# Compute and plot Sensitivity Matrix Method (SMM) results
SMM results are generated from the variational matrices `vareq1` and `vareq2` using the function `calcSensitivityFromMatrix`. The `outputs` argument controls which SMM indicators are computed.
They are ([more details](https://lapp.nl/lapp-software/parid.html)):

- The null space dimension `N`. This is a categorical indicator, equal to the number of unidentifiable directions. So 0 means identifiable, and larger than 0 means unidentifiable. 
- The skewing angle `A`. This is a continuous indicator, taking values between 0 and 1, where 0 means unidentifiable, 1 means identifiable. If `A` is close to 0 then the model is close to unidentifiable.
- The unidentifiable directions `R`. This lists the parameter directions in which the model is unidentifiable. The number of directions equals `N`.
- The minimal parameter relations `M` and the M-norm given by its `norm` attribute. The M-norm is a continuous indicator, taking values between 0 and 1, where 0 means unidentifiable, 1 means identifiable. If the M-norm is close to 0 then the model is close to unidentifiable. For small M-norms, the vector `M` contains the parameter direction in which the model is unidentifiable.
- The least identifiable parameter norms (L-norms) `L`. This is a vector of continuous indicators, one for each parameter, taking values between 0 and 1, where 0 means unidentifiable, 1 means identifiable. If a component of `L` is close to 0 then that parameter is close to unidentifiable.

The results are displayed and graphed.
The first variant, where the bio-availability F was fixed, is identifiable, as can be seen from the output of `simplifySensitivities(sens1)`: the null space dimension `N` is 0, and the skewing angle, M-norm and L-norms are all large.

```{r}
sens1 <- calcSensitivityFromMatrix(outputs = list("N", "A", "R", "M", "L"), df = vareq1, vars = vartheta1)
simplifySensitivities(sens1)
plsens1 <- plotSensitivities(sens1)
plsens1 <- lapply(plsens1[lengths(plsens1) > 0], function(pl) pl + theme_bw(base_size = 8))
cowplot::plot_grid(plotlist = plsens1, nrow = 1)
```

The second variant, where F was variable, is not identifiable, see the output of `simplifySensitivities(sens2)`: the null space dimension is 1, and the skewing angle, M-norm and L-norms are all small or even zero.
The indicators `R` and `M` find a relation involving F, CL and V, as expected.

```{r}
sens2 <- calcSensitivityFromMatrix(outputs = list("N", "A", "R", "M", "L"), df = vareq2, vars = vartheta2)
simplifySensitivities(sens2)
plsens2 <- plotSensitivities(sens2)
plsens2 <- lapply(plsens2[lengths(plsens2) > 0], function(pl) pl + theme_bw(base_size = 8))
cowplot::plot_grid(plotlist = plsens2, nrow = 1)
```

# Compute and plot Aliasing Method results
The results are generated from the variational matrices `vareq1` and `vareq2` using the function `calcAliasingFromMatrix`.
The results are displayed and graphed.
Both variants are identifiable, as aliasing scores are quite low. This is because the aliasing method can only find identifiability problems involving two parameters, and not three.

```{r}
alia1 <- calcAliasingScoresFromMatrix(df = vareq1, vars = vartheta1)
plalia1 <- plotAliasing(alia1, elt = c("S", "T"))
cowplot::plot_grid(plotlist = plalia1, nrow = 1)

alia2 <- calcAliasingScoresFromMatrix(df = vareq2, vars = vartheta2)
plalia2 <- plotAliasing(alia2, elt = c("S", "T"))
cowplot::plot_grid(plotlist = plalia2, nrow = 1)
```

# Compute and plot Fisher Information Matrix Method (FIMM) results
FIMM results are generated from the variational matrices `vareq1` and `vareq2` using the functions `calcFimFromMatrix` and `fimIdent`, that compute the Fisher Information Matrix (FIM) and the parameter identifiability indicators, respectively.
The indicators are ([more details](https://lapp.nl/lapp-software/parid.html)):

- Categorical identifiability `identifiable`. This is `TRUE` if all curvatures are 0 and the model is identifiable, and `FALSE` if there are positive curvatures and the model is unidentifiable. The curvatures describe the objective function value (OFV) surface as function of the parameters.
- The number of 0 curvatures `nDirections`. This is a categorical indicator counting the number of 0 curvatures.
- The parameter vectors `directions` corresponding to the curvatures, in order of increasing curvature. For zero (or small) curvatures, they indicate the directions in parameter space of (near) unidentifiability.
- The curvature values `curvatures` of the OFV surface. Zero values correspond to unidentifiability, small values indicate near unidentifiability.
- The index `jump` of the highest change in curvature value.
- The estimated standard error `se`, calculated from the FIM.
- The estimated relative standard error `rse`, calculated from the FIM.

The results are displayed.
The first variant, where the bio-availability F was fixed, is identifiable: all curvatures are high. The options `relChanges = TRUE` and `ci = 0.95` imply that the `directions` show the relative percentage changes corresponding to an insignificant increase (at 95%) in objective function of at most 3.84, for the given number of subjects. They are all below 50%.

```{r}
fim1 <- calcFimFromMatrix(df = vareq1, omega = omega, sigma = sigma, vartheta = vartheta1)
fimres1 <- fimIdent(fim = fim1, curvature = 1e-10, relChanges = TRUE, ci = 0.95, nsubj = 30)
simplifyFimIdent(fimres1)
```

The second variant, where F was variable, is not identifiable, as the first curvature is close to zero. It finds a relation involving F, CL and V in `directions`, as expected. The relative changes are very large, indicating that the parameters can change by a large factor without significantly changing the OFV.

```{r}
fim2 <- calcFimFromMatrix(df = vareq2, omega = omega, sigma = sigma, vartheta = vartheta2)
fimres2 <- fimIdent(fim = fim2, curvature = 1e-10, relChanges = TRUE, ci = 0.95, nsubj = 30)
simplifyFimIdent(fimres2)
```

